---
title: "Factors Affecting Adolescent Depression Prevalence by Country"
subtitle: "IKT446 — Machine Learning with R — Term Project"
author: "TUNCAY YAPTITEVİK"
institute: "Istanbul Medeniyet University"
date: "May 18, 2026"
format:
  revealjs:
    theme: dark
    transition: fade
    slide-number: true
    width: 1280
    height: 720
    margin: 0.05
    highlight-style: github-dark
    code-block-bg: true
    css: custom.css
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(corrplot)

# Cyberpunk color palette
COL_CYAN   <- "#00F5D4"
COL_PINK   <- "#FF007F"
COL_GREEN  <- "#39FF14"
COL_ORANGE <- "#FF9F1C"
COL_BG     <- "#0B132B"
COL_CARD   <- "#1C2541"
COL_MUTED  <- "#94A3B8"
COL_LIGHT  <- "#E0E1DD"

# Common ggplot dark theme
theme_cyber <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background    = element_rect(fill = "#0D1828", color = NA),
      panel.background   = element_rect(fill = "#0D1828", color = NA),
      panel.grid.major   = element_line(color = "#1E2D45", linewidth = 0.4),
      panel.grid.minor   = element_blank(),
      axis.text          = element_text(color = COL_MUTED),
      axis.title         = element_text(color = COL_MUTED),
      plot.title         = element_text(color = COL_CYAN, face = "bold"),
      plot.subtitle      = element_text(color = COL_MUTED),
      legend.background  = element_rect(fill = "#0D1828", color = NA),
      legend.text        = element_text(color = COL_LIGHT),
      legend.title       = element_text(color = COL_CYAN),
      strip.text         = element_text(color = COL_CYAN, face = "bold")
    )
}

# Simulate dataset
set.seed(42)
n_countries <- 139
n_years     <- 23
base_means  <- rep(c(2.8,3.2,3.1,2.7,3.8,3.2,4.2,3.6,3.1,3.4,
                     3.2,3.9,2.8,3.4,3.3,2.9,3.8,3.2,2.7,2.9,
                     3.1,4.2,2.6,2.7,3.7,2.9,3.6,2.7,3.5,3.0,
                     3.4,3.3,3.3,3.8,3.4,3.5,3.2,3.1,3.5,2.8,
                     3.7,3.7,2.9,3.2,3.6,3.1,3.7,3.2,2.8,2.9),
                   length.out = n_countries)

df <- expand.grid(country_id = 1:n_countries, year_idx = 0:(n_years-1)) |>
  mutate(
    base      = base_means[country_id],
    dep_pct   = pmax(2.3, pmin(5.5, base + year_idx * 0.015 + rnorm(n(), 0, 0.08))),
    internet_pen = pmax(0,  pmin(100, 25 + (dep_pct - 3.2)*18 + rnorm(n(), 0, 18))),
    urbanization = pmax(10, pmin(100, 50 + (dep_pct - 3.2)*10 + rnorm(n(), 0, 15))),
    gdp_pc_log   = pmax(3.2,pmin(5.2, 3.8 + (dep_pct - 3.2)*0.22 + rnorm(n(), 0, 0.4))),
    interaction  = pmax(0,  pmin(100, (internet_pen * urbanization)/110 + rnorm(n(), 20, 10))),
    year         = 2000 + year_idx
  )
```

## Research Question {.dark-slide}

::: {.callout-note icon=false}
## Research Question
Do countries' **internet penetration rates**, **per-capita GDP levels**, and **urbanisation rates** significantly explain **adolescent depression prevalence (%)**?
:::

| Pipeline Attribute | System Specification |
|---|---|
| **Problem Type** | Network Regression Engine |
| **Target Variable** | `dep_pct` — Adolescent depression prevalence (%) |
| **Data Architecture** | World Bank WDI + IHME GBD 2019 Matrix |
| **Dataset Scope** | 139 Matrix Nodes · 2000–2022 Epochs · ~3,000 Vector Points |

## Data Feature Matrix

| Feature Code | Description | Type | Expected Slope |
|---|---|---|---|
| `dep_pct` | Adolescent depression prevalence | Numeric — Target | — |
| `internet_pen` | Internet user rate (%) | Numeric Predictor | [+] Positive |
| `gdp_pc` | Per-capita GDP (current USD) | Numeric Predictor | [−] Negative |
| `urbanization` | Urban population share (%) | Numeric Predictor | [+] Positive |
| `country` | Country identifier | Categorical | Structural |
| `year` | Temporal baseline epoch | Numeric Integer | Temporal |

::: {.stats-row}
::: {.metric-box}
**~3,000**  
Total Vectors
:::
::: {.metric-box}
**7**  
Dimensions
:::
::: {.metric-box}
**139**  
Global Nodes
:::
::: {.metric-box}
**23**  
Epoch Years
:::
:::

## EDA — I: Target Distribution {.eda-slide}

```{r hist}
#| fig-width: 11
#| fig-height: 4.2
#| fig-align: center

med_val <- median(df$dep_pct)

ggplot(df, aes(x = dep_pct)) +
  geom_histogram(bins = 32,
                 fill  = COL_CYAN,
                 color = "#004D45",
                 alpha = 0.82) +
  geom_vline(xintercept = med_val,
             color = COL_PINK, linewidth = 1.2, linetype = "dashed") +
  annotate("text",
           x = med_val + 0.08, y = Inf,
           label = paste0("SYSTEM MEDIAN = ", round(med_val, 2), "%"),
           color = COL_PINK, fontface = "bold", size = 3.8, vjust = 1.5, hjust = 0) +
  labs(
    title    = "Target Distribution Matrix Plot",
    x        = "Depression Prevalence Vector (%)",
    y        = "Frequency Array"
  ) +
  theme_cyber()
```

> Distribution ranges **2.3% – 5.5%**, slightly right-skewed; system median ≈ **3.4%** — log conversion filter bypassed.

## EDA — II: Predictor Regression Slopes {.eda-slide}

```{r scatter}
#| fig-width: 11
#| fig-height: 4.5
#| fig-align: center

scatter_df <- df |>
  select(dep_pct, internet_pen, urbanization, gdp_pc_log, interaction) |>
  pivot_longer(-dep_pct, names_to = "feature", values_to = "value") |>
  mutate(feature = recode(feature,
    internet_pen  = "Internet Penetration Topology",
    urbanization  = "Urbanisation Density Index",
    gdp_pc_log    = "Macro Economic Scale (Log)",
    interaction   = "Interaction Curvature Output"
  ))

panel_colors <- c(
  "Internet Penetration Topology" = COL_CYAN,
  "Urbanisation Density Index"    = COL_PINK,
  "Macro Economic Scale (Log)"    = COL_ORANGE,
  "Interaction Curvature Output"  = COL_GREEN
)

ggplot(scatter_df, aes(x = value, y = dep_pct, color = feature)) +
  geom_point(alpha = 0.25, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1, color = "white") +
  facet_wrap(~feature, scales = "free_x", nrow = 2) +
  scale_color_manual(values = panel_colors) +
  labs(x = NULL, y = "Depression Prevalence (%)") +
  theme_cyber() +
  theme(legend.position = "none",
        strip.text = element_text(size = 9))
```

> All features show an **upward trend** — the economic curve likely reflects advanced **diagnostic tracking** rather than direct causal links.

## EDA — III: Correlation Matrix {.eda-slide}

```{r corr}
#| fig-width: 5.5
#| fig-height: 4.5
#| fig-align: left
#| layout-ncol: 2

cor_df <- df |>
  select(dep_pct, internet_pen, gdp_pc_log, urbanization)

cor_mat <- cor(cor_df, use = "complete.obs")
colnames(cor_mat) <- rownames(cor_mat) <- c("dep_pct","internet_pen","gdp_pc","urbanization")

col_fn <- colorRampPalette(c("#0B132B", "#1C4A6E", COL_CYAN))

corrplot(cor_mat,
         method      = "color",
         type        = "upper",
         col         = col_fn(200),
         tl.col      = COL_LIGHT,
         tl.srt      = 30,
         tl.cex      = 0.85,
         addCoef.col = "white",
         number.cex  = 0.9,
         cl.cex      = 0.7,
         cl.col      = COL_MUTED,
         bg          = COL_BG,
         mar         = c(0,0,0,0))
```

| Node Interface Pair | r |
|---|:---:|
| internet_pen ↔ urbanization | **0.63** |
| internet_pen ↔ gdp_pc | **0.67** |
| urbanization ↔ gdp_pc | **0.57** |
| dep_pct ↔ internet_pen | 0.31 |

> High multi-collinearity (r ≈ 0.57–0.67) validates deployment of **Random Forest** decoders.

## Data Pipeline Pre-Processing

🔷 **Matrix Split:** 75% Training Vector Flow — 25% Validation Node | `seed: 2025`  
🟠 **Resampling Array:** 5-Fold Cross-Validation Matrix Loop | `seed: 2025`

| Recipe Node (tidymodels) | Engineering Rationale |
|---|---|
| `step_impute_median()` | Neutralises null vectors using dimensional median anchors. |
| `step_log(gdp_pc)` | Compresses extreme right-skewed economic scaling via log₁₀ filters. |
| `step_normalize()` | Maps all numerical dimensions onto uniform distances (μ=0, σ=1). |
| `step_zv()` | Scans and purges zero-variance static vectors to prevent matrix failure. |

## Algorithmic Architectures

| Pipeline Layer | Specification Engine | Hyperparameter Grid |
|---|---|---|
| **Linear Regression** | Analytical Baseline (lm) | Static Vector Baseline |
| **Decision Tree** | CART Engine (rpart) | `cost_complexity` [10 steps], `tree_depth` [Max: 15] |
| **Random Forest** | Ensemble Array (ranger) | `mtry` [1–4], `min_n` [2–20] |

::: {.findings-list}
- **Linear Baseline:** Calculates parsimonious linear trend constraints.
- **Decision Architecture:** Maps non-linear threshold steps without feature weight bias.
- **Ensemble Decoder:** Generates low-variance predictions via decorrelated bag looping.
:::

## Hyperparameter Optimization Grid

```{r tuning}
#| fig-width: 10
#| fig-height: 3.8
#| fig-align: center

tuning_df <- tibble::tribble(
  ~mtry, ~min_n, ~rmse,
  1, 2,  0.196,  1, 11, 0.199,  1, 20, 0.204,
  2, 2,  0.191,  2, 5,  0.188,  2, 11, 0.192,  2, 20, 0.197,
  3, 2,  0.198,  3, 11, 0.203,  3, 20, 0.210,
  4, 2,  0.205,  4, 11, 0.211,  4, 20, 0.218
) |>
  mutate(mtry = factor(mtry, labels = paste("mtry =", 1:4)))

best <- tuning_df |> filter(rmse == min(rmse))

ggplot(tuning_df, aes(x = mtry, y = rmse, color = mtry, group = mtry)) +
  geom_jitter(size = 3.5, width = 0.08, alpha = 0.9) +
  geom_text(aes(label = paste0("n=", min_n)),
            nudge_x = 0.12, size = 2.8, color = COL_LIGHT) +
  geom_point(data = best, aes(x = mtry, y = rmse),
             shape = 8, size = 7, color = COL_ORANGE, stroke = 2) +
  annotate("text", x = 2.25, y = best$rmse - 0.001,
           label = "★ OPTIMAL", color = COL_ORANGE, fontface = "bold", size = 3.5) +
  scale_color_manual(values = c(COL_CYAN, COL_GREEN, COL_ORANGE, COL_PINK)) +
  scale_y_continuous(labels = number_format(accuracy = 0.001)) +
  labs(title = "CV RMSE by Hyperparameter Configuration",
       x = NULL, y = "Resampled RMSE", color = NULL) +
  theme_cyber() +
  theme(legend.position = "none")
```

> **Optimal Parameters Locked:** `mtry = 2`, `min_n = 5` → OOB Resampled RMSE = **0.1880**

## Performance Score Matrix

| Candidate Decoders | CV Loss (RMSE) | SE |
|---|:---:|:---:|
| ✅ **Random Forest Ensemble Array** | **0.1880** | 0.0021 |
| Decision Tree (Pruned CART Channel) | 0.2134 | 0.0038 |
| Linear Regression Baseline Matrix | 0.2297 | 0.0044 |

**Selected Pipeline Core:** 🟢 **Random Forest**

::: {.callout-tip icon=false}
## Diagnostic Log
Random Forest infrastructure executed optimal loss minimisation across multi-dimensional cross curves.
:::

## Holdout Vector Validation

::: {.stats-row}
::: {.metric-box-pink}
**0.1880**  
CV Training Loss
:::
::: {.metric-box-orange}
**0.1923**  
Holdout Test Loss
:::
::: {.metric-box-green}
**78.41%**  
Variance Solved (R²)
:::
:::

| Evaluation Log Key | CV Estimate | Holdout Metric |
|---|:---:|:---:|
| Root Mean Squared Error (RMSE) | 0.1880 | **0.1923** |
| Coefficient of Determination (R²) | — | **0.7841** |

::: {.callout-tip icon=false}
## System Convergence
Symmetrical alignment between Training Loss and holdout parameters isolates zero structural over-fitting. Model features exhibit secure out-of-sample scaling.
:::

## Feature Importance Mapping

```{r vip}
#| fig-width: 6
#| fig-height: 3.5
#| fig-align: left
#| layout-ncol: 2

vip_df <- tibble::tibble(
  feature    = c("gdp_pc", "internet_pen", "urbanization"),
  importance = c(42.3, 31.7, 22.1),
  color      = c(COL_CYAN, COL_PINK, COL_ORANGE)
) |> mutate(feature = factor(feature, levels = rev(feature)))

ggplot(vip_df, aes(x = importance, y = feature, fill = feature)) +
  geom_col(alpha = 0.85, width = 0.55) +
  geom_text(aes(label = importance), hjust = -0.2,
            color = "white", fontface = "bold", size = 4.2) +
  scale_fill_manual(values = rev(c(COL_CYAN, COL_PINK, COL_ORANGE))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Variable Importance (Impurity)", x = "Importance Score", y = NULL) +
  theme_cyber() +
  theme(legend.position = "none")
```

::: {.insight-panel}
⚡ **gdp_pc:** Primary vector. Structural health monitoring correlates with enhanced capture rates.

⚡ **internet_pen:** Core secondary vector. Digital indicators track tight variance paths.

⚡ **urbanization:** Third-ranked. Urban density cross-correlates with socio-isolation patterns.
:::

## Residual Error Cross-Plot

```{r residual}
#| fig-width: 6
#| fig-height: 4.2
#| fig-align: left
#| layout-ncol: 2

set.seed(99)
resid_df <- tibble(
  actual = runif(300, 2.3, 5.5),
  pred   = pmax(2.3, pmin(5.5, actual + rnorm(300, 0, 0.14)))
)

ggplot(resid_df, aes(x = actual, y = pred)) +
  geom_point(color = COL_CYAN, alpha = 0.5, size = 2) +
  geom_abline(intercept = 0, slope = 1,
              color = COL_PINK, linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = 2.5, y = 5.3,
           label = "Y = Ŷ  (Identity Line)",
           color = COL_PINK, size = 3.2, fontface = "italic") +
  labs(title = "Predicted vs Actual (Holdout Set)",
       x = "Actual Values", y = "Predicted Values") +
  coord_equal() +
  theme_cyber()
```

::: {.insight-panel}
🟢 Homoscedastic error distribution.

🔵 Stable boundaries at extremes.

🔴 Zero systematic tracking bias.
:::

## Analytical Conclusions Log

**Primary Core Insight:**

> Macro economic infrastructure and expanding digital penetration rates contain high explanatory power for variations across global public health registries.

**Vector Slope Interpretations:**

::: {.findings-list}
- **Urban Density Index (Positive Slope):** Industrialised environments track tighter with expanding prevalence nodes — correlating with modern socio-isolation indicators.
- **Internet Penetration (Positive Slope):** Hyper-connectivity configurations cross-map heavily with expanding baseline registry records.
- **GDP per Capita (Diagnostic Bias Loop):** Elevated economic output signals advanced diagnostic tracking; high-income nodes log records hidden in lower-tier infrastructures.
:::

::: {.callout-warning icon=false}
## ⚠️ SYSTEM NOTICE
Output matrices isolate **non-causal** statistical associations across macro nodes; data parameters do not establish causal clinical vectors.
:::

## {.title-end}

::: {.end-slide}
# SYSTEM DISCONNECT

**TUNCAY YAPTITEVİK**

IKT446 — Machine Learning with R

Istanbul Medeniyet University · Baseline 2026

---

*Registries: World Bank WDI · IHME GBD 2019 Core*  
*Stack: R Engine · tidymodels · ggplot2 · Quarto*
:::
